Fail Faster

### 2025-03-21 paper/paper.tex
methodology

\subsubsection{OCaml}
\label{subsubsection:ocaml}

In \camlname, our strategies include type-derived generators for BSTs and STLC terms
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed'').

The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
using \csplitmix. The single-pass BST generator benefits more from
staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
\csplitmix. We attribute the difference in performance speedup to the fact that
BST strategies rely on relatively few calls to generator combinator functions,
but sample extensively from the \rand; notably, single-pass BST, which uses the largest
number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}

For STLC, staging accounts for the bulk of performance gains, yielding
$1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
the well-typed generator. Adding \csplitmix, these numbers increase to
$2.60-5.15\times$ and $4.83-7.75\times$. This result makes sense because
STLC generators, particularly the well-typed generator, sample very little from
randomness libraries but rely extensively on generator library functions.

@@

\subsubsection{Scala 3}
In \scalaname, we implement a subset of the \camlname
strategies: the boolean list generator, single-pass BST generator, and
well-typed STLC term generator. ScalaCheck does not support type derivation,
so those strategies are excluded.

>>>>

\subsubsection{\camlname}
\label{subsubsection:ocaml}

In \camlname, our strategies include type-derived generators for BSTs and STLC terms, 
following the approach described in \secref{subsection:type-derived}; two custom 
BST generators—--one that builds a tree incrementally by repeatedly inserting 
values into an initially empty structure, and another that constructs the 
entire tree in a single pass by generating keys, values, and subtrees at each step; 
a boolean list generator that mirrors the single-pass BST strategy; and a 
generator for STLC terms that is correct by construction (i.e., it produces 
only well-typed terms).

These strategies differ in their use of generator library combinators 
and in how frequently they sample random values. The type-derived and 
insertion-based BST generators sample frequently but use combinators 
sparingly. The single-pass BST generator also samples frequently 
but uses more combinators, striking a middle ground. 
The type-derived 
STLC generator samples infrequently and makes moderate 
use of combinators, while the custom STLC generator uses 
significantly more combinators than any other strategy and 
also takes the fewest samples. The boolean list generator 
makes moderate use of combinators 
and samples booleans frequently---however, because booleans are 
inexpensive to sample, its overall sampling overhead is lower 
than that of the other strategies.

For each strategy, we compare three treatments: our \bq baseline; 
a \camlname version using \bq's randomness 
library, \texttt{splittable\_random}; and a
\camlname version using \csplitmix. For each treatment, we measure the
time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
variety of different sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc. Regardless of how size is encoded, our goal is to show that performance trends scale.}
We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.

Our results are summarized in \figref{fig:time}.
Our type-derived \camlname BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
using \csplitmix. The single-pass BST generator benefits more from
staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
\csplitmix. 
% We attribute the difference in performance speedup to the fact that
% BST strategies rely on relatively few calls to generator combinator functions,
% but sample extensively from the \rand; notably, single-pass BST, which uses the largest
% number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}

For STLC, staging accounts for the bulk of performance gains, yielding
$1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
the well-typed generator. Adding \csplitmix, these numbers increase to
$2.60-5.15\times$ and $4.83-7.75\times$.

% This result makes sense because
% STLC generators, particularly the well-typed generator, sample very little from
% randomness libraries but rely extensively on generator library functions.

@@

\subsubsection{\scalaname}
In \scalaname, we implement a subset of the \camlname
strategies: the boolean list generator, single-pass BST generator, and
well-typed STLC term generator. ScalaCheck does not support type derivation,
so those strategies are excluded.
